---
title: "mirBottleneck Analysis Report"
date: "`r format(Sys.time(), '%d %B %Y, %H:%M')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    self_contained: true
    code_folding: hide
params:
  results:     NULL
  cohort_name: "TCGA-PAAD"
  pkg_version: "0.99.0"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, fig.align="center")
library(plotly); library(DT); library(survival)
res          <- params$results
combined     <- res$combined_bottleneck
cox_ind      <- res$cox_individual
cox_final    <- res$cox_final
survival_df  <- res$survival_df
targets      <- res$mirna_targets
clinical     <- res$clinical_df
vss_scores   <- res$vss_scores
coherence_scores <- res$coherence_scores
patient_scores   <- res$patient_scores
contributing     <- res$contributing_mirnas
class_colors <- c(dual="#D55E00", silencer="#0072B2", conductor="#009E73", weak="#999999")
```

<style>
body{font-family:'Segoe UI',Arial,sans-serif;font-size:14px}
.section-box{background:#f8f9fa;border-left:4px solid #2E4B7A;padding:12px 16px;border-radius:4px;margin-bottom:16px}
.metric-grid{display:grid;grid-template-columns:repeat(auto-fit,minmax(160px,1fr));gap:12px;margin-bottom:20px}
.metric-card{background:white;border:1px solid #dee2e6;border-radius:8px;padding:16px;text-align:center;box-shadow:0 1px 3px rgba(0,0,0,.08)}
.metric-value{font-size:28px;font-weight:700;color:#2E4B7A}
.metric-label{font-size:11px;color:#6c757d;text-transform:uppercase;letter-spacing:.5px;margin-top:4px}
h2{color:#2E4B7A;border-bottom:2px solid #e9ecef;padding-bottom:6px}
h3{color:#495057}
</style>

<div class="section-box">
**Package:** mirBottleneck v`r params$pkg_version` &nbsp;|&nbsp;
**Cohort:** `r params$cohort_name` &nbsp;|&nbsp;
**Generated:** `r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`
</div>

## 1. Cohort Summary

```{r cohort-metrics}
n_patients <- nrow(clinical)
n_events   <- sum(clinical$OS_status, na.rm=TRUE)
med_age    <- median(clinical$age, na.rm=TRUE)
n_surv     <- nrow(survival_df)
```

<div class="metric-grid">
<div class="metric-card"><div class="metric-value">`r n_patients`</div><div class="metric-label">Total Patients</div></div>
<div class="metric-card"><div class="metric-value">`r n_surv`</div><div class="metric-label">With Survival Data</div></div>
<div class="metric-card"><div class="metric-value">`r n_events`</div><div class="metric-label">Events (Deaths)</div></div>
<div class="metric-card"><div class="metric-value">`r med_age`</div><div class="metric-label">Median Age (yrs)</div></div>
<div class="metric-card"><div class="metric-value">`r nrow(combined)`</div><div class="metric-label">miRNAs Scored</div></div>
<div class="metric-card"><div class="metric-value">`r nrow(contributing)`</div><div class="metric-label">Contributing miRNAs</div></div>
</div>

```{r stage-plot, fig.height=3}
if (!is.null(clinical$stage_clean)) {
  stg <- as.data.frame(table(Stage=clinical$stage_clean))
  stg <- stg[!is.na(stg$Stage), ]
  plot_ly(stg, x=~Stage, y=~Freq, type="bar",
    marker=list(color="#2E4B7A"),
    hovertemplate="%{x}: %{y} patients<extra></extra>") |>
  layout(title="Tumour Stage Distribution",
    xaxis=list(title="Stage"), yaxis=list(title="N"),
    plot_bgcolor="white", paper_bgcolor="white")
}
```

## 2. miRNA–Target Network

```{r network-metrics}
n_mirna  <- nrow(targets)
med_tgts <- median(targets$n_targets)
max_tgts <- max(targets$n_targets)
```

<div class="metric-grid">
<div class="metric-card"><div class="metric-value">`r n_mirna`</div><div class="metric-label">miRNAs with Targets</div></div>
<div class="metric-card"><div class="metric-value">`r med_tgts`</div><div class="metric-label">Median Targets / miRNA</div></div>
<div class="metric-card"><div class="metric-value">`r max_tgts`</div><div class="metric-label">Max Targets</div></div>
</div>

```{r network-plot, fig.height=3.5}
top_net <- head(targets[order(-targets$n_targets), ], 15)
top_net$label <- gsub("hsa-mir-", "miR-", top_net$mirna)
plot_ly(top_net, x=~reorder(label, n_targets), y=~n_targets, type="bar",
  marker=list(color="#0072B2"),
  hovertemplate="%{x}: %{y} targets<extra></extra>") |>
layout(title="Top 15 miRNAs by Validated Target Count",
  xaxis=list(title="", tickangle=-35), yaxis=list(title="N validated targets"),
  plot_bgcolor="white", paper_bgcolor="white")
```

## 3. Variance Suppression Score (VSS)

```{r vss-plot, fig.height=3.5}
plot_ly(x=~vss_scores$vss, type="histogram", nbinsx=20,
  marker=list(color="#2E4B7A", line=list(color="white", width=0.5)),
  hovertemplate="VSS: %{x:.4f}<br>Count: %{y}<extra></extra>") |>
layout(title="VSS Distribution",
  xaxis=list(title="Variance Suppression Score"), yaxis=list(title="Count"),
  shapes=list(list(type="line",
    x0=median(vss_scores$vss), x1=median(vss_scores$vss),
    y0=0, y1=1, yref="paper",
    line=list(color="#D55E00", dash="dash", width=2))),
  plot_bgcolor="white", paper_bgcolor="white")
```

```{r vss-table}
top_vss <- head(vss_scores[order(-vss_scores$vss), ], 10)
top_vss$mirna <- gsub("hsa-mir-", "hsa-miR-", top_vss$mirna)
top_vss$vss   <- round(top_vss$vss, 5)
datatable(top_vss, caption="Top 10 miRNAs by VSS", rownames=FALSE,
  options=list(pageLength=10, dom="t"),
  colnames=c("miRNA", "VSS", "N Targets"))
```

## 4. Coherence Induction Score

```{r coherence-plot, fig.height=3.5}
sig_n <- sum(combined$coherence_p < 0.05, na.rm=TRUE)
plot_ly(x=~combined$coherence_score, type="histogram", nbinsx=20,
  marker=list(color="#009E73", line=list(color="white", width=0.5)),
  hovertemplate="Coherence: %{x:.4f}<br>Count: %{y}<extra></extra>") |>
layout(title=paste0("Coherence Score Distribution  (", sig_n, " miRNAs p < 0.05)"),
  xaxis=list(title="Coherence Induction Score"), yaxis=list(title="Count"),
  shapes=list(list(type="line", x0=0, x1=0, y0=0, y1=1, yref="paper",
    line=list(color="#D55E00", dash="dash", width=2))),
  plot_bgcolor="white", paper_bgcolor="white")
```

## 5. Bottleneck Classification

```{r phase-diagram, fig.height=5}
vss_med <- median(combined$vss_norm)
coh_med <- median(combined$coherence_norm)
combined$label <- gsub("hsa-mir-", "miR-", combined$mirna)
combined$sz    <- scales::rescale(combined$bottleneck_index, to=c(5, 18))

plot_ly(combined,
  x=~vss_norm, y=~coherence_norm,
  type="scatter", mode="markers",
  color=~class, colors=class_colors,
  marker=list(size=~sz, opacity=0.85, line=list(color="white", width=0.5)),
  text=~paste0("<b>", label, "</b><br>Class: ", class,
    "<br>VSS: ", round(vss, 5),
    "<br>Coherence: ", round(coherence_score, 4),
    "<br>Bottleneck Index: ", round(bottleneck_index, 4)),
  hoverinfo="text") |>
layout(
  title="miRNA Bottleneck Classification — Phase Diagram",
  xaxis=list(title="VSS (normalised)", zeroline=FALSE),
  yaxis=list(title="Coherence Score (normalised)", zeroline=FALSE),
  shapes=list(
    list(type="line", x0=vss_med, x1=vss_med, y0=0, y1=1, yref="paper",
      line=list(color="grey", dash="dot", width=1.5)),
    list(type="line", y0=coh_med, y1=coh_med, x0=0, x1=1, xref="paper",
      line=list(color="grey", dash="dot", width=1.5))),
  annotations=list(
    list(x=0.98, y=0.98, text="<b>Dual</b>",      xref="paper", yref="paper", showarrow=FALSE, font=list(color="#D55E00", size=11), xanchor="right"),
    list(x=0.98, y=0.02, text="<b>Silencer</b>",  xref="paper", yref="paper", showarrow=FALSE, font=list(color="#0072B2", size=11), xanchor="right"),
    list(x=0.02, y=0.98, text="<b>Conductor</b>", xref="paper", yref="paper", showarrow=FALSE, font=list(color="#009E73", size=11), xanchor="left"),
    list(x=0.02, y=0.02, text="<b>Weak</b>",      xref="paper", yref="paper", showarrow=FALSE, font=list(color="#999",    size=11), xanchor="left")),
  plot_bgcolor="white", paper_bgcolor="white",
  legend=list(title=list(text="<b>Archetype</b>")))
```

```{r archetype-bar, fig.height=3}
cls_tbl <- as.data.frame(table(Class=combined$class))
cls_tbl$color <- class_colors[as.character(cls_tbl$Class)]
cls_tbl$Class <- tools::toTitleCase(as.character(cls_tbl$Class))
plot_ly(cls_tbl, x=~Class, y=~Freq, type="bar",
  marker=list(color=~color),
  hovertemplate="%{x}: %{y} miRNAs<extra></extra>") |>
layout(title="Archetype Distribution",
  xaxis=list(title=""), yaxis=list(title="N miRNAs"),
  plot_bgcolor="white", paper_bgcolor="white")
```

## 6. Top Bottleneck miRNAs

```{r top-table}
cox_sub <- cox_ind[, c("mirna","hr","p")]
mrg     <- merge(combined, cox_sub, by="mirna", all.x=TRUE)
mrg     <- mrg[order(-mrg$bottleneck_index), ]
out <- data.frame(
  miRNA            = gsub("hsa-mir-", "hsa-miR-", mrg$mirna),
  Class            = tools::toTitleCase(mrg$class),
  VSS              = round(mrg$vss, 5),
  Coherence        = round(mrg$coherence_score, 4),
  `Bottleneck Index`= round(mrg$bottleneck_index, 4),
  `HR (OS)`        = round(mrg$hr, 3),
  `p-value`        = signif(mrg$p, 3),
  check.names=FALSE)
datatable(out, caption="All scored miRNAs ranked by bottleneck index",
  rownames=FALSE, filter="top",
  options=list(pageLength=15, scrollX=TRUE)) |>
formatStyle("p-value",
  color=styleInterval(0.05, c("#D55E00","#333")),
  fontWeight=styleInterval(0.05, c("bold","normal")))
```

## 7. Survival Analysis

```{r survival-metrics}
cox_s  <- summary(cox_final)
hr_val <- round(cox_s$coefficients["composite","exp(coef)"], 3)
ci_lo  <- round(cox_s$conf.int["composite", 3], 3)
ci_hi  <- round(cox_s$conf.int["composite", 4], 3)
p_val  <- signif(cox_s$coefficients["composite","Pr(>|z|)"], 3)
c_idx  <- round(cox_s$concordance[1], 3)
lr     <- survdiff(Surv(OS_days, OS_status) ~ score_group, data=survival_df)
lr_p   <- signif(pchisq(lr$chisq, df=1, lower.tail=FALSE), 3)
med_hi <- median(survival_df$OS_days[survival_df$score_group=="High"], na.rm=TRUE)
med_lo <- median(survival_df$OS_days[survival_df$score_group=="Low"],  na.rm=TRUE)
```

<div class="metric-grid">
<div class="metric-card"><div class="metric-value">`r hr_val`</div><div class="metric-label">Hazard Ratio</div></div>
<div class="metric-card"><div class="metric-value">`r paste0(ci_lo,"–",ci_hi)`</div><div class="metric-label">95% CI</div></div>
<div class="metric-card"><div class="metric-value">`r p_val`</div><div class="metric-label">Cox p-value</div></div>
<div class="metric-card"><div class="metric-value">`r c_idx`</div><div class="metric-label">C-index</div></div>
<div class="metric-card"><div class="metric-value">`r lr_p`</div><div class="metric-label">Log-rank p</div></div>
<div class="metric-card"><div class="metric-value">`r med_hi` vs `r med_lo`</div><div class="metric-label">Median OS (days) Hi vs Lo</div></div>
</div>

```{r km-plot, fig.height=5}
km    <- survfit(Surv(OS_days, OS_status) ~ score_group, data=survival_df)
km_df <- data.frame(time=km$time, surv=km$surv, lower=km$lower, upper=km$upper,
  group=rep(names(km$strata), km$strata))
km_df$grp <- ifelse(grepl("High", km_df$group), "High Score", "Low Score")

plot_ly() |>
add_trace(data=km_df[km_df$grp=="High Score",],
  x=~time, y=~surv, type="scatter", mode="lines", name="High Score",
  line=list(color="#D55E00", width=2.5),
  hovertemplate="Day %{x}<br>Survival: %{y:.3f}<extra>High Score</extra>") |>
add_trace(data=km_df[km_df$grp=="High Score",],
  x=~c(time,rev(time)), y=~c(upper,rev(lower)),
  type="scatter", mode="none", fill="toself",
  fillcolor="rgba(213,94,0,0.1)", showlegend=FALSE, hoverinfo="skip") |>
add_trace(data=km_df[km_df$grp=="Low Score",],
  x=~time, y=~surv, type="scatter", mode="lines", name="Low Score",
  line=list(color="#0072B2", width=2.5),
  hovertemplate="Day %{x}<br>Survival: %{y:.3f}<extra>Low Score</extra>") |>
add_trace(data=km_df[km_df$grp=="Low Score",],
  x=~c(time,rev(time)), y=~c(upper,rev(lower)),
  type="scatter", mode="none", fill="toself",
  fillcolor="rgba(0,114,178,0.1)", showlegend=FALSE, hoverinfo="skip") |>
layout(
  title=paste0("Kaplan-Meier Overall Survival  (log-rank p = ", lr_p, ")"),
  xaxis=list(title="Time (days)", zeroline=FALSE),
  yaxis=list(title="Survival Probability", range=c(0,1), zeroline=FALSE),
  plot_bgcolor="white", paper_bgcolor="white",
  legend=list(x=0.75, y=0.95),
  annotations=list(list(
    x=0.02, y=0.05, xref="paper", yref="paper", showarrow=FALSE, align="left",
    text=paste0("HR = ",hr_val," (",ci_lo,"–",ci_hi,")<br>Cox p = ",p_val,"  |  C-index = ",c_idx),
    font=list(size=11, color="#333"),
    bgcolor="#f8f9fa", bordercolor="#dee2e6", borderwidth=1)))
```

```{r cox-table}
cc <- as.data.frame(summary(cox_final)$coefficients)
raw_vars <- rownames(cc)
var_labels <- raw_vars
var_labels[var_labels == "composite"]      <- "Composite Score"
var_labels[var_labels == "age"]            <- "Age"
var_labels[grepl("stage", var_labels)]     <- paste0("Stage ", gsub("stage_clean","", var_labels[grepl("stage", var_labels)]))
cc$Variable <- var_labels
cc <- cc[, c("Variable","exp(coef)","se(coef)","z","Pr(>|z|)")]
colnames(cc) <- c("Variable","HR","SE","z","p-value")
cc$HR <- round(cc$HR,3); cc$SE <- round(cc$SE,4)
cc$z  <- round(cc$z,3);  cc$`p-value` <- signif(cc$`p-value`,3)
datatable(cc, rownames=FALSE, options=list(dom="t",pageLength=10)) |>
formatStyle("p-value",
  color=styleInterval(0.05, c("#D55E00","#333")),
  fontWeight=styleInterval(0.05, c("bold","normal")))
```

## 8. Per-Patient Composite Score

```{r patient-scores, fig.height=4}
sc_df <- data.frame(patient=names(patient_scores), score=as.numeric(patient_scores))
sc_df <- sc_df[order(sc_df$score), ]
sc_df$color <- ifelse(sc_df$score >= median(sc_df$score), "#D55E00", "#0072B2")
sc_df$group <- ifelse(sc_df$score >= median(sc_df$score), "High", "Low")

plot_ly(sc_df, x=~seq_along(score), y=~score, type="bar",
  marker=list(color=~color),
  text=~paste0(patient,"<br>Score: ",round(score,4),"<br>Group: ",group),
  hoverinfo="text") |>
layout(
  title="Composite Bottleneck Score per Patient",
  xaxis=list(title="Patients (ranked)", showticklabels=FALSE),
  yaxis=list(title="Composite Score"),
  shapes=list(list(type="line",
    x0=0, x1=nrow(sc_df), y0=median(sc_df$score), y1=median(sc_df$score),
    line=list(color="#333", dash="dash", width=1.5))),
  plot_bgcolor="white", paper_bgcolor="white")
```

## 9. Methods Summary

<div class="section-box">

**Pipeline:** mirBottleneck v`r params$pkg_version`

**Cohort:** `r params$cohort_name` — `r nrow(clinical)` patients (`r nrow(survival_df)` with survival data)

**Network:** `r nrow(targets)` miRNAs from miRTarBase via multiMiR; minimum 5 validated targets required.

**VSS:** Mean R² from linear regression of each validated target on miRNA expression across patients.

**Coherence:** Δ mean pairwise Pearson correlation between high- and low-expression patient groups; empirical p-value by permutation (n = 200).

**Classification:** Median splits of normalised VSS and coherence scores → dual / silencer / conductor / weak.

**Composite score:** Direction-aware weighted mean of miRNA expression for contributing miRNAs (individual Cox p < 0.15); weights = |Cox coefficient|.

**Survival:** Cox proportional hazards adjusted for age + tumour stage; log-rank test for High vs Low groups.

</div>

## 10. Session Info

```{r session, echo=TRUE}
sessionInfo()
```

---

<div style="text-align:center;color:#aaa;font-size:11px;margin-top:30px">
Generated by mirBottleneck v`r params$pkg_version` &nbsp;·&nbsp;
<a href="https://github.com/MarkBarsoumMarkarian/mirBottleneck" style="color:#aaa">GitHub</a> &nbsp;·&nbsp;
<a href="https://doi.org/10.5281/zenodo.19121116" style="color:#aaa">Zenodo</a>
</div>
