---
title: "mirBottleneck: miRNA Bottleneck Scoring Workflow"
author: "Mark Barsoum Markarian"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{mirBottleneck: miRNA Bottleneck Scoring Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  message   = FALSE,
  warning   = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

`mirBottleneck` identifies miRNAs that act as **transcriptome stabilizers** in
cancer — miRNAs whose expression suppresses variance and induces coordinated
programs across their validated target genes. It scores each miRNA along two
complementary axes:

- **Variance Suppression Score (VSS):** how much a miRNA reduces target gene
  expression variance across patients.
- **Coherence Score:** how much a miRNA coordinates its targets into coherent
  co-expression programs (Δ mean pairwise correlation, high vs. low expressors).

miRNAs are classified into four archetypes (**dual**, **silencer**,
**conductor**, **weak**) and a patient-level **composite bottleneck index** is
derived from the top survival-associated bottleneck miRNAs. The index is
evaluated against overall survival using Cox proportional hazards models.

This vignette demonstrates the full workflow using the **small synthetic toy
dataset** shipped with the package (10 patients, 50 genes, 30 miRNAs). No
network access or large downloads are required. For a real TCGA-PAAD run, see
the `run_mirBottleneck_project()` section at the end.

# Installation

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install("mirBottleneck")
```

# Load the package

```{r load}
library(mirBottleneck)
```

# Toy dataset

The package ships five small synthetic objects in `inst/extdata/`. Use
`.toy_paths()` to locate them — no hard-coded paths needed.

```{r toy-paths}
paths <- mirBottleneck:::.toy_paths()
str(paths)
```

```{r load-toy}
mirna_log      <- readRDS(paths$mirna_log)       # 30 miRNAs x 10 patients
rna_sym        <- readRDS(paths$rna_sym)          # 50 genes  x 10 patients
clinical_df    <- readRDS(paths$clinical_df)      # 10 patients x 9 cols
mirna_norm_map <- readRDS(paths$mirna_norm_map)   # 30-row map
mirna_targets  <- readRDS(paths$mirna_targets)    # precomputed toy network

cat("miRNA matrix :", nrow(mirna_log), "miRNAs x", ncol(mirna_log), "patients\n")
cat("RNA matrix   :", nrow(rna_sym),   "genes  x", ncol(rna_sym),   "patients\n")
cat("Clinical     :", nrow(clinical_df), "patients,", ncol(clinical_df), "variables\n")
```

```{r show-clinical}
head(clinical_df)
```

# Step 1 — miRNA–target network

In a real analysis, `build_network()` queries miRTarBase via `multiMiR` to
fetch validated interactions. Here we use the **precomputed toy network** so
the vignette runs offline and deterministically.

```{r show-network}
head(mirna_targets[, c("mirna", "n_targets")])
```

```{r build-network-real, eval=FALSE}
# Real usage (requires internet + multiMiR):
mirna_targets <- build_network(
  mirna_ids   = rownames(mirna_log),
  rna_symbols = rownames(rna_sym)
)
```

# Step 2 — Variance Suppression Score (VSS)

For each miRNA, `score_vss()` fits a linear model of each validated target
gene on the miRNA expression across patients. VSS = mean R² across all targets.

```{r score-vss}
vss_scores <- score_vss(
  mirna_log      = mirna_log,
  rna_sym        = rna_sym,
  mirna_targets  = mirna_targets,
  mirna_norm_map = mirna_norm_map
)

head(vss_scores)
```

```{r vss-plot, fig.cap="VSS distribution across toy miRNAs"}
hist(
  vss_scores$vss,
  breaks = 10,
  main   = "Variance Suppression Score distribution",
  xlab   = "VSS (mean R² across validated targets)",
  col    = "#4C72B0",
  border = "white"
)
```

# Step 3 — Coherence Induction Score

`score_coherence()` splits patients by median miRNA expression and measures
whether target genes become more correlated in the high-expression group.
An empirical p-value is computed via permutation.

```{r score-coherence}
coherence_scores <- score_coherence(
  mirna_log      = mirna_log,
  rna_sym        = rna_sym,
  mirna_targets  = mirna_targets,
  mirna_norm_map = mirna_norm_map,
  scored_mirnas  = vss_scores$mirna,
  n_perm         = 50,      # low for vignette speed; use 200+ in real analysis
  max_targets    = 50
)

head(coherence_scores)
```

```{r coherence-plot, fig.cap="Coherence score vs. VSS for toy miRNAs"}
merged <- merge(vss_scores, coherence_scores, by = "mirna")
plot(
  merged$vss, merged$coherence_score,
  xlab = "Variance Suppression Score (VSS)",
  ylab = "Coherence Induction Score",
  main = "VSS vs. Coherence — toy dataset",
  pch  = 19,
  col  = "#4C72B0"
)
abline(v   = median(merged$vss),             lty = 2, col = "grey50")
abline(h   = median(merged$coherence_score), lty = 2, col = "grey50")
legend("topright",
       legend = c("median splits"),
       lty = 2, col = "grey50", bty = "n")
```

# Step 4 — Classification

`classify_bottleneck()` normalises both scores to [0, 1], computes a composite
bottleneck index, and assigns each miRNA to one of four archetypes based on
median splits of the normalized scores.

| Class | VSS | Coherence |
|---|---|---|
| **dual** | ≥ median | ≥ median |
| **silencer** | ≥ median | < median |
| **conductor** | < median | ≥ median |
| **weak** | < median | < median |

```{r classify}
classified <- classify_bottleneck(vss_scores, coherence_scores)

table(classified$class)
```

```{r classify-plot, fig.cap="miRNA classification by bottleneck archetype"}
class_cols <- c(
  dual      = "#D55E00",
  silencer  = "#0072B2",
  conductor = "#009E73",
  weak      = "#999999"
)
barplot(
  table(classified$class),
  col    = class_cols[names(table(classified$class))],
  main   = "miRNA bottleneck class distribution",
  ylab   = "Number of miRNAs",
  border = "white"
)
```

# Step 5 — Patient-level composite score

`composite_score()` builds a direction-aware weighted score per patient from
the miRNAs most associated with survival (nominally p < threshold). Weights
are |Cox coefficient|; miRNAs with HR > 1 push the score up.

```{r composite}
composite <- composite_score(
  mirna_log       = mirna_log,
  classified_df   = classified,
  clinical_df     = clinical_df,
  mirna_norm_map  = mirna_norm_map,
  cox_p_threshold = 0.5    # relaxed for toy (only 10 patients)
)

cat("Patients scored:", length(composite$patient_scores), "\n")
cat("Contributing miRNAs:", nrow(composite$contributing_mirnas), "\n")
```

```{r composite-plot, fig.cap="Patient composite bottleneck score distribution"}
scores <- composite$patient_scores
barplot(
  sort(scores),
  main   = "Patient composite bottleneck scores",
  ylab   = "Score",
  col    = ifelse(sort(scores) >= median(scores), "#D55E00", "#0072B2"),
  border = "white",
  las    = 2,
  cex.names = 0.7
)
abline(h = median(scores), lty = 2, col = "grey40")
legend("topleft",
       legend = c("High", "Low"),
       fill   = c("#D55E00", "#0072B2"),
       bty    = "n")
```

# Step 6 — Survival analysis

`survival_model()` fits three Cox models (score alone, clinical alone,
combined) and a log-rank test comparing High vs. Low score groups.

> **Note:** With only 10 toy patients, survival statistics are not
> interpretable — this step is shown for API demonstration only.
> See the [full dataset section](#full-dataset) for a real analysis.

```{r survival}
surv <- survival_model(
  patient_scores = composite$patient_scores,
  clinical_df    = clinical_df
)

cat("Log-rank p-value :", round(surv$logrank_p, 4), "\n")
cat("C-index (score)  :", round(surv$summary["c_score"],    4), "\n")
cat("C-index (combined):", round(surv$summary["c_combined"], 4), "\n")
```

# One-step pipeline with `run_mirBottleneck_project()`

All six steps above are wrapped in a single convenience function. Supply paths
to `.rds` files and an output directory:

```{r pipeline-toy, eval=FALSE}
paths <- mirBottleneck:::.toy_paths()
out   <- file.path(tempdir(), "mirBottleneck_toy_run")

results <- run_mirBottleneck_project(
  mirna_log_rds      = paths$mirna_log,
  rna_rds            = paths$rna_sym,
  clinical_rds       = paths$clinical_df,
  mirna_norm_map_rds = paths$mirna_norm_map,
  mirna_targets_rds  = paths$mirna_targets,  # offline: skip network query
  query_network      = FALSE,
  out_dir            = out,
  n_perm             = 50,
  cox_p_threshold    = 0.5
)

list.files(out)
```

# Full TCGA-PAAD dataset {#full-dataset}

The complete harmonized TCGA-PAAD dataset (~51 MB, 178 patients, 60,660 genes,
1,881 miRNAs) is archived on Zenodo:

**DOI:** https://doi.org/10.5281/zenodo.19121116

```{r fetch-full, eval=FALSE}
# Download once — cached to tools::R_user_dir("mirBottleneck", "cache")
paths_full <- fetch_mirBottleneck_data()

results_full <- run_mirBottleneck_project(
  mirna_log_rds      = paths_full[["mirna_log"]],
  rna_rds            = paths_full[["rna_final"]],
  clinical_rds       = paths_full[["clinical"]],
  mirna_norm_map_rds = paths_full[["mirna_norm_map"]],
  query_network      = TRUE,   # builds network from miRTarBase
  out_dir            = "mirBottleneck_results/"
)
```

# Session info

```{r session}
sessionInfo()
```
